---
title: "clustRviz Quick Start"
author:
 - name: Michael Weylandt
   affiliation: Department of Statistics, Rice University
   email: michael.weylandt@rice.edu
 - name: John Nagorski
   affiliation: Department of Statistics, Rice University
 - name: Genevera I. Allen
   affiliation: |
     | Departments of Statistics, Computer Science, and Electical and Computer Engineering, Rice University
     | Jan and Dan Duncan Neurological Research Institute, Baylor College of Medicine
   email: gallen@rice.edu
date: "Last Updated: August 19th, 2020"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
bibliography: vignettes.bib
vignette: >
  %\VignetteIndexEntry{clustRviz Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval=TRUE,
  message = FALSE
)
```
\renewcommand{\vec}[1]{\boldsymbol{#1}}

## Introduction

This vignette provides a brief introduction to the `clustRviz` package, 
describing how to use the main entry points `CARP` and `CBASS` and providing
a quick overview of the rich built-in graphics functionality. For more details
on graphics, weight selection, or the computational algorithms used, please
see the other package vignettes. 

## Clustering

`clustRviz` implements the *convex* clustering formulation popularized by
Hocking *et al.* [-@Hocking:2011] and uses the path-wise algorithms of
Weylandt, Nagorski, and Allen [-@Weylandt:2019] to support full path
computation and dendrogram construction. This allows convex clustering to 
produce `hclust`-style dendrograms while maintaining its statistical and computational
advantages. 

The main entry point for clustering is the `CARP` function, which implements
the **Clustering via Algorithmic Regularization Paths** proposed by 
Weylandt, Nagorski, and Allen [-@Weylandt:2019]. We can use it on the built-in
`presidential_speech` data set: 

```{r}
library(clustRviz)
carp_fit <- CARP(presidential_speech)
print(carp_fit)
```

As can be seen, this provides a full path in only a few seconds. From this, 
we can produce a variety of attractive plots, including dendrograms

```{r}
plot(carp_fit, type = "dendrogram")
```

one-way heatmaps

```{r}
plot(carp_fit, type = "heatmap")
```

and regularization paths

```{r}
plot(carp_fit, type = "path")
```

For each plot type, interactive and dynamic versions are also supported: for example,

```{r}
plot(carp_fit, type = "dendrogram", dynamic = TRUE)
```

By default, the entire path is shown, but it is possible to obtain specific solutions
by specifying the `k` or `percent` arguments to plot. 

```{r}
plot(carp_fit, k = 3)
```

To work with the clustering solutions directly, the `get_cluster_labels`, `get_clustered_data`, 
or `get_cluster_centroids` functions may be useful. 

## Bi-Clustering

Chi *et al* [-@Chi:2017] proposed a convex formulation of *biclustering* for which
Weylandt [-@Weylandt:2019b] later proposed an efficient ADMM algorithm. This ADMM
was adapted into the **CBASS** - *Convex Biclustering via Algorithmic Regularization
with Small Steps* algorithm. `clustRviz` exposes an implementation of this algorithm
via the function of the same name. 

```{r}
library(clustRviz)
cbass_fit <- CBASS(presidential_speech)
print(cbass_fit)
```

As can be seen, this provides a full path in only a few seconds. In general, the 
bi-clustering problem is a bit slower than the standard clustering problem but still
highly efficient. From this, we can produce a variety of attractive plots, 
including row- and column-wise dendrograms

```{r}
plot(cbass_fit, type = "row.dendrogram")
```

```{r}
plot(cbass_fit, type = "col.dendrogram")
```

row- and columnwise regularization paths

```{r}
plot(cbass_fit, type = "row.path")
```

and the traditional two-way cluster heatmap

```{r}
plot(cbass_fit, type = "heatmap")
```

As before, interactive and dynamic versions are also supported: for example,

```{r}
plot(cbass_fit, type = "heatmap", dynamic = TRUE)
```

Because `CBASS` clusters rows and columns simultaneously, when specifying 
cluster numbers, it is necessary to distinguish between row and column clusters

```{r}
plot(cbass_fit, k.row = 3)
```

This is only a brief demonstration of the capabilities of the `clustRviz` package
- see the other vignettes for more!

## References
