---
title: "clustRviz Quick Start"
author: "John Nagorski"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    css: style.css
vignette: >
  %\VignetteIndexEntry{clustRviz Details}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval=TRUE,
  message = FALSE
)
```
\renewcommand{\vec}[1]{\boldsymbol{#1}}

# Introduction

This vignette provides a quick usage guide to the `clustRviz` package,
including fitting and visualization.
In the sections below we work through a simple example dataset,
illustrating both clustering and biclustering in `clustRviz`.
While many details and customizations are discussed in the
[clustRviz Details Vignette](clustRvizDetails.html),
for now we show how to quickly fit, visualize, and
save in `clustRviz`.

# A Clustering Example

`clustRviz` may be used for clustering together observations based on
similarities among their features.
`clustRviz` accomplishes this by solving the Convex Clustering problem
quickly using the technique of Algorithmic Regularization.
The resulting algorithm, Convex Clustering via Algorithmic Regularization Paths
(or CARP) provides a continuous path of clustering solutions which may
be visualized and interactively explored.

For our first example we use the `petrol` data set from the `MASS` package.

```{r, headdat}
# optionally install MASS package
# install.package('MASS')
data('petrol',package='MASS')
head(petrol)
```

The dataset contains the yield (`Y`) of a petroleum refining process and
with measurements for specific gravity (`SG`), vapour pressure (`VP`),
crude volatility (`V10`),
desired gasoline volatility (`EP`), as well as sample
identification labels (`No`); see `?petrol`for details.

Before clustering our data, we extract the columns of interest
(here `SG`, `VP`, `V10`, and `EP`), and create the input matrix:

```{r}
# create numeric input matrix
Xdat <- as.matrix(petrol[,c('SG','VP','V10','EP')])
head(Xdat)
```

### Fitting with CARP

Clustering is done via a simple call to the `CARP` function below.

```{r}
library(clustRviz)
carp.fit <- CARP(Xdat, labels = petrol$No, X.scale = TRUE)
```

Here we specify the sample number (`No`) as the labels for our observations.
This is done by providing the `labels`. If no `labels` were provided, `CARP` would
attempt to use the rownames of the data matrix, and if no rownames were found, would
automatically generate names.
Additionally, since our variables are measured on different scales, we
set the `X.scale` option to `TRUE` as well; see `?carp` for additional
fitting options as well as examples in the
[clustRviz Details Vignette](clustRvizDetails.html)

Once fitted, a brief description of the fit and parameter choices may be shown:

```{r,message=TRUE}
carp.fit
```

Visualizing the solution path is readily done via the `plot` function.
A traditional dendrogram is displayed by default:

```{r,fig.width=5,fig.height=5,fig.align='center'}
plot(carp.fit)
```

A more detailed visualization of the solution path can be obtained by
plotting the principal components of the cluster means:

```{r,fig.width=5,fig.height=5,fig.align='center'}
plot(carp.fit,type = 'path',percent = .10)
```

The plot above displays the first two principal components of the
data, with black points representing the original data, and red
curves illustrating movement along the solution path. As curves
join together, observations are fused, and clusters formed.
The `percent` argument tells `plot` to display a snapshot of the
clustering solution path at $10\%$ of total regularization.

Both the dendrogram and the clustering solution path may be
viewed together interactively via the use of `shiny` applications:

```{r,eval=FALSE}
plot(carp.fit,type = 'interactive')
```

These interactive visualizations allow one to view the cluster
solution path and dendrogram as the Algorithmic Regularization
iterates vary.

In addition to visualizations, cluster solutions may be obtained
via the `get_cluster_labels` and `get_cluster_centroids` functions:

```{r}
# Analyze the 3 cluster solution
get_cluster_labels(carp.fit, k = 3)

get_cluster_centroids(carp.fit, k = 3)
```

In the code above, we specify the number of clusters and `clustering`
returns the both the cluster assignment as well as the cluster means
for the $k=3$ solution along the path. As an example, we display the
yield (`Y`) distribution for each of the clusters:

```{r, fig.width=5, fig.height=5, fig.align='center'}
library(ggplot2)
ggplot(
  aes(x=cl,y=yield),
  data=data.frame(
    yield = petrol$Y,
    cl=get_cluster_labels(carp.fit, k = 3)
  )
) + geom_boxplot()
```

After exploring visualizations and cluster solutions interactively,
the visualization results can easily be exported via the `saveviz`
function. For example, saving the cluster solution path plotted
above may be done as follows:

```{r,eval=FALSE}
saveviz(
  x=carp.fit,
  file.name="ClusterPathStatic.png",
  plot.type='path',
  image.type = 'static',
  percent=.10)
```

The above saves a static snapshot of the cluster solution path 
at $10\%$ regularization; the number of clusters `k` may
also be specified instead of `percent`.
Interactive visualization may be saved as `.gif` files by
specifiying `image.type='dynamic'`; see `?saveviz.CARP` for
details.

# A Biclustering Example

Clustering may also be done simultaneously on both observations and
variables. `clustRviz` obtains such biclustering solutions by
approximating the Convex Biclustering solution path, again via
Algorithmic Regularization. The `CBASS` function is used to obtain
biclustering solutions, and its usage is similar to `CARP`.

We fit the biclustering solution path as follows:

```{r}
cbass.fit <- CBASS(X=Xdat, row_labels = petrol$No)
```

Once fitted, vizualizations are displayed again via the `plot` function.
For CBASS fits, the default visualization is again the cluster dendrogram
of the observations induced by the CBASS solution path.

```{r,fig.width=5,fig.height=5,fig.align='center'}
plot(cbass.fit)
```

Since both rows (observations) and columns (variables) are clustered simultaneously,
a cluster dendrogram for the columns may also be plotted:

```{r, eval=FALSE}
plot(cbass.fit, type='col.dendrogram')
```

Additionally both dendrograms and the associated cluster heatmap 
can be viewed by passing the `heatmap` arguement.

```{r,fig.width=10,fig.height=10,fig.align='center'}
plot(cbass.fit, type='heatmap')
```

As was the case of `CARP`, interactive visualizations may also be displayed:

```{r,eval=FALSE}
plot(cbass.fit,type='interactive')
```

`CBASS`'s interactive visualizations allows one to view dynamic changes in the
cluster heatmap and dendrogram solutions along the path.

As was the case with `CARP`, clustering solutions are obtained via the
`get_cluster_labels` function. The desired clustering solution is specified by:

- the number of row clusters (`k.row`);
- the number of col clusters (`k.col`); or
- the percent of regularization (`percent`).

In the example below we obtain the biclustering solution
which obtains $3$ row clusters:

```{r}
# How are the rows (observations) clustered at the 3 row cluster solution?
table(get_cluster_labels(cbass.fit, k.row = 3))

# How are the columns (features) clustered at the 3 row cluster solution?
table(get_cluster_labels(cbass.fit, k.row = 3, type = "col"))
```

We can also investigate the bi-clustered data matrix (with the raw data replaced
by estimated centroids) with the `get_clustered_data` function:

```{r}
get_clustered_data(cbass.fit, k.row = 3)
```

Finally we demonstrate how to save the visualizations
produced by `CBASS`. Similar to `CARP` visualizations
can be saved via the `saveviz` function which allows for saving
static and dynamic visuals. For example, a `.gif`
of the variable dendrogram as it varies throughout the
path may be saved via:

```{r,eval=FALSE}
saveviz(x=cbass.fit,
        file.name = 'cbass_column_dendrogram_dynamic.gif',
        plot.type = 'col.dendrogram',
        image.type = 'dynamic'
)
```

Similarly, a static snapshot of the bicluster heatmap at the moment
along the path when $4$ row clusters are formed may be
saved also:

```{r,eval=FALSE}
saveviz(x=cbass.fit,
        file.name = 'cbass_heatmap_static.png',
        plot.type = 'heatmap',
        image.type = 'static',
        k.row = 4
)
```
