---
title: "clustRviz Quick Start"
author: "John Nagorski"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    css: style.css
vignette: >
  %\VignetteIndexEntry{clustRviz Details}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval=TRUE,
  message = FALSE
)
```
\renewcommand{\vec}[1]{\boldsymbol{#1}}

# Introduction

This vignette provides a quick usage guide to the `clustRviz` package, 
including fitting and visualization.
In the sections below we work through a simple example dataset, 
illustrating both clustering and biclustering in `clustRviz`.
While many details and customizations are discussed in the 
[clustRviz Details Vignette](clustRvizDetails.html), 
for now we show how to quickly fit, visualize, and 
save in `clustRviz`. 

# A Clustering Example

`clustRviz` may be used for clustering together observations based on 
similarities among their features.
`clustRviz` accomplishes this by solving the Convex Clustering problem 
quickly using the technique of Algorithmic Regularization.
The resulting algorithm, Convex Clustering via Algorithmic Regularization Paths 
(or CARP) provides a continuous path of clustering solutions which may 
be visualized and interactively explored.

For our first example we use the `petrol` data set from the `MASS` package.

```{r, headdat}
# optionally install MASS package
# install.package('MASS')
data('petrol',package='MASS')
head(petrol)
```

The dataset contains the yield (`Y`) of a petroleum refining process and 
with measurements for specific gravity (`SG`), vapour pressure (`VP`), 
crude volatility (`V10`), 
desired gasoline volatility (`EP`), as well as sample 
identification labels (`No`); see `?petrol`for details. 

Before clustering our data, we extract the columns of interest 
(here `SG`, `VP`, `V10`, and `EP`), and create the input matrix:

```{r}
# create numeric input matrix
Xdat <- as.matrix(petrol[,c('SG','VP','V10','EP')])
head(Xdat)
```

### Fitting with CARP

Clustering is done via a simple call to the `CARP` function below.

```{r}
library(clustRviz)
carp.fit <- CARP(
  Xdat,
  control=list(
    obs.labels = petrol$No,
    X.scale=TRUE)
)
```

Here we specify the sample number (`No`) as the labels for our observations.
This is done by providing `obs.labels` to the `control` argument. 
While not necessary, providing labels adds clarity to many visualizations. 
Additionally, since our variables are measured on different scales, we 
set the `X.scale` option to `TRUE` as well; see `?carp.control` for additional 
fitting options as well as examples in the 
[clustRviz Details Vignette](clustRvizDetails.html)

Once fitted, a brief description of the fit and parameter choices 
may be shown

```{r,message=TRUE}
carp.fit
```

Visualizing the solution path is readily done via the `plot` function.
A traditional dendrogram is displayed by default:

```{r,fig.width=5,fig.height=5,fig.align='center'}
plot(carp.fit)
```

A more detailed visualization of the solution path can be obtained by
plotting the principal components of the cluster means:

```{r,fig.width=5,fig.height=5,fig.align='center'}
plot(carp.fit,type = 'path',percent = .10)
```

The plot above displays the first two principal components of the 
data, with black points representing the original data, and red 
curves illustrating movement along the solution path. As curves 
join together, observations are fused, and clusters formed.
The `percent` argument tells `plot` to display a snapshot of the 
clustering solution path at $10\%$ of total regularization.

Both the dendrogram and the clustering solution path may be 
viewed together interactively via the use of `shiny` applications:

```{r,eval=FALSE}
plot(carp.fit,type = 'interactive')
```

These interactive visualizations allow one to view the cluster 
solution path and dendrogram as the Algorithmic Regularization 
iterates vary.

In addition to visualizations, cluster solutions may be obtained 
via the `clustering` function:

```{r}
# get the 3 cluster solution along the path
carp.clusters <- clustering(carp.fit,k = 3)
# display the labels
table(carp.clusters$clustering.assignment)
# display the cluster mean
head(carp.clusters$cluster.means)
```

In the code above, we specify the number of clusters and `clustering`
returns the both the cluster assignment as well as the cluster means 
for the $k=3$ solution along the path. As an example, we display the 
yield (`Y`) distribution for each of the clusters: 

```{r,fig.width=5,fig.height=5,fig.align='center'}
library(ggplot2)
ggplot(
  aes(x=cl,y=yield),
  data=data.frame(
    yield = petrol$Y,
    cl=carp.clusters$clustering.assignment
  )
) + geom_boxplot()
```

After exploring visualizations and cluster solutions interactively, 
the visualization results can easily be exported via the `saveviz`
function. For example, saving the cluster solution path plotted 
above may be done as follows:

```{r,eval=FALSE}
saveviz(
  x=carp.fit,
  file.name="ClusterPathStatic.png",
  plot.type='path',
  image.type = 'static',
  percent=.10)
```

The above saves a static snapshot of the cluster solution path 
at $10\%$ regularization; the number of clusters `k` may
also be specified instead of `percent`.
Interactive visualization may be saved as `.gif` files by 
specifiying `image.type='dynamic'`; see `?saveviz.CARP` for 
details.

# A Biclustering Example

Clustering may also be done simultaneously on both observations and 
variables. `clustRviz` obtains such biclustering solutions by 
approximating the Convex Biclustering solution path, again via 
Algorithmic Regularization. The `CBASS` function is used to obtain 
biclustering solutions, and its usage is similar to `CARP`.

We fit the biclustering solution path as follows:

```{r}
CBASS(
  X=Xdat,
  control=list(
    obs.labels = petrol$No
  )) -> cbass.fit
```

Once fitted, vizualizations are displayed again via the `plot` function. 
For CBASS fits, the default visualization is again the cluster dendrogram 
of the observations induced by the CBASS solution path.

```{r,fig.width=5,fig.height=5,fig.align='center'}
plot(cbass.fit)
```

Since both observations and variables are clustered simultaneously, a 
cluster dendrogram for the variables may also be plotted:

```{r,eval=FALSE}
plot(cbass.fit,type='var.dendrogram')
```

Additionally both dendrograms and the associated cluster heatmap 
can be viewed by passing the `heatmap` arguement.

```{r,fig.width=10,fig.height=10,fig.align='center'}
plot(cbass.fit,type='heatmap')
```

As was the case of `CARP`, interactive visualizations may also be displayed:

```{r,eval=FALSE}
plot(cbass.fit,type='interactive')
```

`CBASS`'s interactive visualizations allows one to view dynamic changes in the 
cluster heatmap and dendrogram solutions along the path.

As was the case with `CARP`, clustering solutions are obtained via the 
`clustering` function. The desired clustering solution is returned by
specifying either the number of observation clusters (`k.obs`), 
the number of variable clusters (`k.var`), or the percent of 
regularization (`percent`). In any case, `clustering` returns 
the observation and variable clustering assignments, as well as
the matrix of cluster means.

In the example below we obtain the biclustering solution 
which obtains $3$ observation clusters:

```{r}
# obtain the desired clustering solution along the path
cbass.clusters <- clustering(cbass.fit,k.obs = 3)
# observation clusters
table(cbass.clusters$clustering.assignment.obs)
# variable clusters
table(cbass.clusters$clustering.assignment.var)
# mean matrix
cbass.clusters$cluster.mean.matrix[1:5,]
```

Finally we demonstrate how to save the visualizations 
produced by `CBASS`. Similar to `CARP` visualizations 
can be saved via the `saveviz` function which allows for saving 
static and dynamic visuals. For example, a `.gif` 
of the variable dendrogram as it varies throughout the 
path may be saved via:

```{r,eval=FALSE}
saveviz(x=cbass.fit,
        file.name = 'cbass_vardend_dynamic.png',
        plot.type = 'var.dendrogram',
        image.type = 'dynamic'
)
```

Similarly, a static snapshot of the bicluster heatmap at the moment 
along the path when $4$ observation clusters are formed may be 
saved also:
```{r,eval=FALSE}
saveviz(x=cbass.fit,
        file.name = 'cbass_heatmap_static_kobs.png',
        plot.type = 'heatmap',
        image.type = 'static',
        k.obs = 4
)
```
