---
title: "10-Minute Introduction to ClustRviz"
author:
 - name: Michael Weylandt
   affiliation: Department of Statistics, Rice University
   email: michael.weylandt@rice.edu
 - name: John Nagorski
   affiliation: Department of Statistics, Rice University
 - name: Genevera I. Allen
   affiliation: |
     | Departments of Statistics, Computer Science, and Electical and Computer Engineering, Rice University
     | Jan and Dan Duncan Neurological Research Institute, Baylor College of Medicine
   email: gallen@rice.edu
date: "Last Updated: August 19th, 2020"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
bibliography: vignettes.bib
vignette: >
  %\VignetteIndexEntry{10-Minute Introduction to clustRviz}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval=TRUE,
  message = FALSE
)
```

\renewcommand{\vec}[1]{\boldsymbol{#1}}

# Introduction

The `clustRviz` package intends to make fitting and visualizing CARP and CBASS
solution paths an easy process. In the [QuickStart](clustRviz.html)
vignette we provide a quick start guide for basic usage, fitting,
and plotting. In this vignette, we build on the basics and provide a more
detailed explanation for the variety of options available in `clustRviz`.

# Background

The starting point for CARP is the Convex Clustering
[@Hocking:2011; @Chi:2015; @Tan:2015] problem:

\[\text{arg min}_{U} \frac{1}{2}\|U - X\|_F^2 + \lambda\sum_{(i, j) \in \mathcal{E}} w_{ij} \|U_{i\cdot} - U_{j\cdot}\|_q\]

where $X$ is an $n \times p$ data matrix, consisting of $p$ measurements on $n$ subjects,
$\lambda \in \mathbb{R}_{> 0}$ a regularization parameter determining the degree of clustering, and
$w_{ij}>0$ a weight for each pair of observations; here $\| . \|_F$ and $\| . \|_2$ denote
the Frobenius norm and $\ell_q$ norm, respectively. (Typically, we take $q = 2$.)

Briefly, Convex Clustering seeks to find and estimate a matrix $\hat{U} \in \mathbb{R}^{n\times p}$
which is both faithful to the original data (Frobenius norm loss term) and has columns
"shrunk" together by the fusion penalty term. The penalty term induces "clusters"
of equal columns in $U$ - if the difference is zero, then the columns must be equal -
while still maintaining closeness to the original data. The equal columns of $U$
can then be used to assign clusters among the columns of $X$.

Unlike other clustering parameters, $\lambda$ can be varied smoothly. At one end,
for zero or very small $\lambda$, the fusion penalty has minimal effect and all
the columns of $\hat{U}$ are distinct, implying $n$ different (trivial) clusters.
At the other end, for large $\lambda$, all columns of $U$ are fused together into
a single "mono-cluster." Between these extremes, every other degree of clustering
can be found as $\lambda$ varies. Considered as a function of $\lambda$,
the columns of $\hat{U}_{\lambda}$ form continuous paths, which we will visualize
below.

To solve the Convex Clustering problem, Chi and Lange [-@Chi:2015] proposed
to use the Alternating Direction Method of Multipliers (ADMM) [@Boyd:2011], an
iterative optimization scheme. This performs well for a single value of $\lambda$,
but can become very expensive if we wish to investigate enough values of $\lambda$
to form a (near-continuous) solution path. To address this, we adapt the Algorithmic
Regularization framework proposed by Hu *et al.* [-@Hu:2016]: this framework takes
a single ADMM step, followed by a small increase in $\lambda$. Remarkably, if the increases
in $\lambda$ are sufficiently small, it turns out that we can approximate the
*true* solution path in a fraction of the time needed for full optimization. Additionally,
because we consider a fine grid of $\lambda$ values, we are able to get a much more
accurate sense of the structure of the solution paths. This approach is called
**Clustering via Algorithmic Regularization Paths** or **CARP** and is implemented
in the function of the same name.

This tutorial will focus on convex *clustering* though the `clustRviz` package also
provides functionality for convex *biclustering* via the **Convex Biclustering via
Algorithmic Regularization with Small Steps** or **CBASS** algorithm, which combines
algorithmic regularization with the Generalized ADMM proposed by Weylandt [-@Weylandt:2019].
All of the graphics and data manipulation functionality demonstrated below for
the `CARP` function work equally well for the `CBASS` function.


## Data Preparation and Weight Calculation

While the `CARP` and `CBASS` functions provides several reasonable default choices
for weights, algorithms, etc, it is important to know their details if one wishes
to compute more customized clustering choices. Here we examine several of the
inputs to `CARP` in more detail.

Here we use a dataset of presidential speechs obtained from
[The American Presidency Project](http://www.presidency.ucsb.edu/index_docs.php)
to illustrate the use of `clustRviz`.
The presidential speech data set contains the top 75 most variable
log-transformed word counts of each US president, aggregated over several
speeches. Additional text processing such as removing stop words and
stemming have been done via the `tm` package.

Let's begin by loading our package and the dataset:

```{r}
library(clustRviz)
data("presidential_speech")
presidential_speech[1:5, 1:5]
```

As can be seen above, this data already comes to us as a *data matrix* with
row and column labels. This is the best format to provide data to `CARP`, though
if labels are missing, default labels will be automatically created.

## Pre-Processing

An important first choice before clustering is whether to center and scale our observations.
Column-wise centering is typically appropriate and is done by default, though this can be changed
with the `X.center` argument. The matter of scaling is a bit more delicate and depends
on whether the features are in some way directly comparable, in which case the raw
scale may be meaningful, or are fundamentally incompatible, in which case it may
make sense to remove scale effects. Scaling is controlled by the `X.scale` argument
to `CARP` and is not implemented for `CBASS`. If scaling is requested, it is performed
by the `scale` function from base `R` which performs unbiased ($n-1$) scaling.
In the case of the presidental speech dataset, all variables are of the same type and
so we will not scale our data matrix.

### Clustering Weights

The choice of clustering weights $w_{ij}$ is essential to getting reasonable
clustering performance. `CARP` uses a reasonably robust heuristic that
balances statistical and computational performance, but great improvements
can be obtained by using bespoke weighting schemes. `CARP` allows weights to be
specified as either a function which returns a weight matrix or a matrix of
weights. More details are available in the [Weights vignette](Weights.html)

### Dimension Reduction and Feature Selection

Convex clustering, as implemented in `clustRviz`, relies on the Euclidean distance
(Frobenius norm) between points. As such, it is rather sensitive to "noisy" features.
To achieve good performance in high-dimensional (big $p$) problems, it is often necessary
to aggressively filter features or perform some sort of dimension reduction before
clustering. Future versions of `clustRviz` will implement the sparse convex
clustering method of Wang *et al* [-@Wang:2018] and the `GeCCO+` scheme
of Wang and Allen [-@Wang:2019], both of which allow for automatic feature weighting
and selection.

`clustRviz` plots leading principal components by default in visualizations, though
this is configurable. Note that the principal components projection is performed
at fit time, not at plot time, so if higher principal components are desired, it
may be necessary to increase the `npcs` argument to `CARP.`

## Fitting

`clustRviz` aims to make it easy to compute the *CARP* and *CBASS* solution paths, and
to quickly begin exploring the results. For now, we will use the defaults described
above to cluster the presidential speech data via the `CARP` function:

```{r, message = FALSE}
carp_fit <- CARP(presidential_speech)
```

Once completed, we can examine a brief summary of the fitted object:

```{r}
carp_fit
```

The default output gives several useful pieces of information, including:

- the sample size ($n = 44$) and number of features in our data ($p = 75$);
- the amount of time necessary to fit the cluster path (approximately 4 seconds);
- the preprocessing performed (centering, but not scaling); and
- the weighting scheme used: 
    - Radial Basis Function Kernel Weights based on the Euclidean Distance
    - A scale parameter $\phi = 0.01$ selected to maximize weight informativeness
    - A high degree of sparsification (four nearest neighbors) to improve computation.

Each of these can be changed using parameters already discussed. Additionally, the
printed output shows the step-size parameter $t$ (default, used here, of 1.05). This
means that at each iteration, $\lambda$ is increased by 5%. For finer path approximations,
the user can use a smaller value of $t$, though be warned that this increases both
computation time and memory usage.

`CARP` has three additional flags that control optimizer behavior:

- `norm` (default `2`): should an $\ell_2$ or $\ell_1$ fusion penalty be used?
  The $\ell_2$ penalty is rotationally symmetric and plays well with downstream
  principal components visualization, so it is used by default.
- `exact` (default `FALSE`): if this is set, the algorithmic regularization scheme
  is replaced with an exact optimization approach. This is often much more expensive,
  but guarantees exact solutions when they matter.
- `back_track` (default `FALSE`): in almost all settings, the monotonic increase
  of $\lambda$ is sufficient to recover the structure of the convex clustering
  dendrogram. For pathological cases where the dendrogram is not easily recovered,
  enabling back-tracking will cause `CARP` to use a back-tracking bisection search
  to precisely identify when observations are fused together.

More advanced optimizer controls can be set using the `clustRviz_options` function.

## Accessing Clustering Solutions

Once fit, the clustering solution may be examined via
three related "accessor" functions:

- `get_cluster_labels`: to get a named factor vector of cluster labels
- `get_cluster_centroids`: to get a matrix of cluster centroids
- `get_clustered_data`: to get the clustered data matrix
  (data replaced by estimated centroids)

The interface for these functions is essentially the same for `CARP` and `CBASS`
objects, though the exact meaning of "centroids" varies between the problems
(vectors for `CARP` and scalars for `CBASS`). The latter two functions also support
a `refit` flag, which determines whether the convex clustering centroids or the naive
centroids (based only on the convex clustering labels) are returned.

For example, we can extract the clustering labels from our `carp_fit` corresponding
to a $k = 2$ cluster solution:

```{r}
cluster_labels <- get_cluster_labels(carp_fit, k = 2)
head(cluster_labels)
```

We see a rather inbalanced data set (the "pre-WWII" cluster is much larger):

```{r}
table(cluster_labels)
```

Similarly, to get the cluster means, we use the `get_cluster_centroids` function:

```{r}
get_cluster_centroids(carp_fit, k = 2)
```

Since we performed convex clustering here, our centroids are $p$-vectors. By default,
the naive centroids are used; if we prefer the exact convex clustering solution, we
can pass the `refit = FALSE` flag:

```{r}
get_cluster_centroids(carp_fit, k = 2, refit = FALSE)
```

We can instead supply the `percent` argument to specify $\lambda$ (or more precisely,
$\lambda / \lambda_{\text{max}}$) rather than the numer of clusters explicitly. For
example, if we are interested at the clustering solution about $25\%$ of the way
along the regularization path:

```{r}
get_cluster_labels(carp_fit, percent = 0.25)
```

We see that our data is clearly falls into three clusters.

Simiarly to `CARP` objects, `CBASS` clustering solutions may also be extracted via the
three accessor functions. The `CBASS` methods allow one of three parameters to be
used to specify the solution:

- `k.row`: the number of row clusters
- `k.col`: the number of column clusters
- `percent`: the percent of total regularization

Other than this, the behavior of `get_cluster_labels`, and `get_clustered_data`
is roughly the same:

```{r, eval = FALSE}
# CBASS Cluster Labels for rows (observations = default)
get_cluster_labels(cbass_fit, percent = 0.85, type = "row")

# CBASS Cluster Labels for columns (features)
get_cluster_labels(cbass_fit, percent = 0.85, type = "col")

# CBASS Solution - naive centroids
get_clustered_data(cbass_fit, percent = 0.85)

# CBASS Solution - convex bi-clustering centroids
get_clustered_data(cbass_fit, percent = 0.85, refit = FALSE)
```

The `get_cluster_centroids` function returns a $k_1$-by-$k_2$ matrix, giving
the (scalar) centroids at a solution with $k_1$ row clusters and $k_2$
column clusters:

```{r, eval = FALSE}
get_cluster_centroids(cbass_fit, percent = 0.85)
```

## Visualizations

`clustRviz` provides a rich set of visualization tools based on the `ggplot2`
and `plotly` libraries. Because `clustRviz` integrates with these libraries,
it is easy to develop custom visualizations based on `clustRviz` defaults.

### Path Plots

The first type of path supported by `clustRviz` is one obtained by plotting values
of $\hat{U}_{\lambda}$ for various values of $\lambda$. Because $\hat{U}$ is a continuous
function of $\lambda$, these paths are continuous and make for attractive visualizations.
As mentioned above, `clustRviz` defaults to plotting the first two principal components
of the data: 

```{r}
plot(carp_fit, type = "path")
```

By default, the entire path is shown, but we can specify a number of clusters directly: 

```{r}
plot(carp_fit, type = "path", k = 3)
```

These plots are standard `ggplot` objects and can be customized in the usual manner: 

```{r}
plot(carp_fit, type = "path", k = 3) + ggplot2::theme_bw() + 
  ggplot2::ggtitle("My Title", subtitle = "Very custom!")
```

We can also plot the raw features directly: 

```{r}
plot(carp_fit, type = "path", axis = c("british", "soviet"))
```

In this space, clear differences in pre-Cold War and post-Cold War topics of presidential
speeches are evident. This sort of customized plot can be saved to a file using
the standard `ggplot2::ggsave` functionality. 

Because paths are continuous, it is also possible to set the `dynamic = TRUE` flag
to produce a GIF via the `gganimate` package: 

```{r}
plot(carp_fit, type = "path", dynamic = TRUE)
```

Finally, interactive path plots can be produced using the `plotly` library
by setting `interactive = TRUE`: 

```{r}
plot(carp_fit, type = "path", dynamic = TRUE, interactive = TRUE)
```

These are currently rather slow to render; we hope
to improve rendering performance in future versions of `clustRviz`.

### Dendrograms

No discussion of clustering visualization would be complete without dendrograms!
`clustRviz` computes a dendrogram based on the order that observations are fused,
with the height of the tree corresponding to the value of $\lambda$ at which the
fusion was first detected. As before, both static and dynamic dendrograms are supported

```{r}
plot(carp_fit, type = "dendrogram")
```

```{r}
plot(carp_fit, type = "dendrogram", dynamic = TRUE)
```

If we specify the number of clusters, a "cut line" is added and the dendrogram
is colored according to cluster identity:

```{r}
plot(carp_fit, type = "dendrogram", k = 3)
```

As before, this is a `ggplot` object and can be manipulated as such. If the underlying
dendrogram object is needed, it can be accessed directly with the `as.dendrogram`
or `as.hclust` functions, allowing for tight integration with other clustering
functionality:

```{r}
as.dendrogram(carp_fit)
as.hclust(carp_fit)
```

As before, an interactive version is also supported

```{r}
plot(carp_fit, type = "dendrogram", interactive = TRUE)
```

### Heatmaps

Finally, a convex clustering-based cluster heatmap is also supported. Cluster heatmaps
are more common for biclustering and we show that here:

```{r}
cbass_fit <- CBASS(presidential_speech)
plot(cbass_fit, type = "heatmap")
```

Unlike the classical cluster heatmap, the convex bi-clustering heatmap is formed
by simultaneously clustering rows and columns, rather than clustering them independently.
This allows us to "cut" the row and column heatmaps at a directly comparable value
of $\lambda$:

```{r}
plot(cbass_fit, type = "heatmap", percent = 0.25)
```

The dynamic plot is particularly compelling:

```{r}
plot(cbass_fit, type = "heatmap", dynamic = TRUE)
```

An interactive heatmap is provided, with help from the excellent `heatmaply` package,
though performance is somewhat limited on larger data sets, so we omit it here.

## Discussion

`clustRviz` provides a integrated framework for fitting and visualizing the *CARP*
and *CBASS* solution paths. `clustRviz` delivers fast computation relative
traditional Convex Clustering solution techniques, and brings traditional and
modern clustering visualization techniques together in a unified framework.

## References
